Curve counting in genus one: \\ elliptic singularities {\it \&} relative geometry

### 2020-11-09 Response to referee/AG Genus One Response to Referee.tex

\def\bibfont{\footnotesize}

\begin{document}

\title{Response to referee report: AG 608 (Version 1)}
%\author{}
\maketitle

\noindent We thank the referee for their careful reading of our paper, and their many valuable suggestions which have greatly helped in improving the quality of the exposition.

We have made many improvements to the paper which directly address the referee's comments. Following their suggestion, we have also gone through and made additional changes, in order to improve the readability and appeal to a wider audience. Major such changes include:
\begin{enumerate}
	\item An improved introduction (see \S\S 1.1--1.5), explicating the connection between logarithmic structures and moduli of Gorenstein singularities which is a key geometric ingredient in the paper.
	\item More details in the proof of Theorem 1.9 (logarithmic smoothness of the moduli space), including a careful analysis of the various relative deformation/obstruction theories. 
\end{enumerate}

@@

\subsection*{Response to specific comments}
\begin{enumerate}
\item[(25)] At the start of Section 5.1 we have added formulae to clarify the definition of the invariants.
\item[(26)] In Section 5.2 we define ``true tangency'' and ``true markings''; we have including more motivation for considering true and fictitious markings. In Section 5.3 we have added (following the table) a more detailed explanation of the recursion algorithm, which clarifies why all the base cases have degree zero. We have moreover discussed these cases in more depth (see Remark 5.4), clarifying the role played by ``tangency'' in this degenerate setting.
\item[(27)] We have clarified the meaning of $\alpha-e_1$. Here true markings can have $\alpha_1=1$, as long as all insertions are not pulled back along forgetting $x_1$; we have added a remark to this effect in Section 5.2. The term ``reduced relative'' is potentially confusing and has been removed.
\item[(28)] Throughout Section 5 we have replaced the ambiguous phrase ``is recursively known'' by the more precise ``has been computed earlier in the recursion'' or ``can be expressed as a polynomial in previously computed invariants''.
\item[(29)] Throughout Section 5 we have replaced this phrase with an explicit formula.
\item[(30)]
\item[(31)] We point out that this uses the same arguments as in Step~1. In the Step~1 arguments, we add explicit references to Gathmann's genus zero results and the necessary recursions on genus one Deligne--Mumford space.
\item[(32)] We have broken up Step 2 and Step 4 into smaller pieces. At several points we have made use of numbered lists for case analysis, which clarifies the logical structure. We have also provided summaries within steps as well as at the end of each step, to help the reader follow which parts of the formula the present arguments pertain to.
\end{enumerate}

\subsection*{Additional minor changes}
\begin{enumerate}
\item The term ``special rubber'' used at the start of Step 2 is an anachronism from a draft version of the paper, and has been removed.
\end{enumerate}

### 2020-11-09 EllipticSingularitiesSplitting.tex
Finished Navid changes

A key step in our recursion is a description of the locus of maps with higher than prescribed tangency, which was identified by Gathmann. We realise the locus as the vanishing locus of a section of a line bundle that comes from tropical geometry — from a piecewise linear function on the fan. The systematic understanding of logarithmic line bundles arising in this fashion is likely to play an important role in logarithmic enumerative geometry.

@@

Our analysis leads naturally to the study of the main component of the double ramification cycle for target manifolds in genus one. The virtual geometry of this was studied in recent work of Janda, Pandharipande, Pixton, and Zvonkine~\cite{DRCBundle}. Even in genus one, an understanding of the main component contributions of the double ramification cycle for target manifolds seems to be lacking. Our main algorithm subverts the use of this class, but it would be more natural to use it.

@@

The moduli stack $\mathfrak M_{1,n}^{\mathrm{cen}}$  of centrally aligned curves is a logarithmic algebraic stack in the smooth topology, and $\mathfrak M_{1,n}^{\mathrm{cen}}\to\mathfrak M_{1,n}$ is a logarithmic modification. The main construction of~\cite{RSPW} canonically associates to any centrally aligned family of curves $\mathcal C_S$ a partial destabilisation $\widetilde{\mathcal C}_S$, introducing $2$-pointed components, and a contraction
\[
\widetilde{\mathcal C}_S\to \overline{\mathcal C}_S,
\]
where $\overline{\mathcal C}_S$ may contain a Gorenstein elliptic singularity \cite{SMY1}. Given a family of tropical curves, the function $\delta$ provides a non-negative real number for each point on the base. We may therefore consider the ``\textbf{circle}'' of points on each fibre whose distance from the core is equal to $\delta$. A partial destabilization, or ``bubbling'', introduces $2$-pointed rational curves into the curve by adding vertices to the dual graph at the points that lie on this circle (the crosses at height $\delta$ in Figure \ref{fig:lambdaexample}); we denote this new tropical curve by $\widetilde{\plC}$. The open disc of radius $\delta$, i.e. the locus where $\lambda<\delta$, is then contracted to a Gorenstein singularity in $\overline{\mathcal C}_S$. The number of branches is equal to the number of vertices of $\widetilde{\sqC}$ which lie on the circle. %This data uniquely determines the singularity.

@@

The first  factor can be described very concretely as the choice of a point $p$ of $H$, and $d$ lines through it. A very important observation is that, at least generically, the vertices which lie on the circle of radius $\delta$ and map to $0\in\mathbb R_{\geq 0}$ are automatically aligned. Indeed, if $R_i$ denotes the $i$-th tail of the curve, glued to $E$ at the node $q_i$, then the stable map provides us with identifications:
\[\operatorname{d}\! f_{i,q_i}\colon T_{R_i,q_i}\cong N_{H/\PP^m,p}\]
which we may compose to obtain attaching data $\theta_{ij}$  (see \S \ref{S:ellsing}). This produces a contraction $C\to\overline C$ to an elliptic $d$-fold point. The dimension of $\widetilde\Dcal$ coincides with the dimension of $\Ecal$, which is $dm + m - 1$.

Assume for simplicity that $d\leq m$. For the absolute factorisation to hold, the $d$ lines $f_i(R_i)$ must span a subspace of $T_{\PP^m,p}$ of dimension at most $d-1$. The locus of maps satisfying this condition has dimension:
\begin{equation*} (dm+m-1) - (m-d+1) = dm + d -2.\end{equation*}
Without loss of generality, we may assume that $H=\{z_0=0\}$, $p=[0,\ldots,0,1]$, and that the $d$ lines span the subspace $\{z_{d-1}=\ldots=z_{m-1}=0\}$. The sections $z_i$ for $i=d-1,\ldots,m$ vanish identically near $E$, so descend automatically to the singularity. On the other hand the section $z_0$ descends to the singularity by the construction of $\overline C$. Finally, the sections $z_i$ for $i=1,\ldots,d-2$ impose independent conditions to descend to $\overline C$. Imposing these conditions, we find that $\Dcal$ has dimension 
\begin{equation*} (dm+d-2)-(d-2) = dm \end{equation*}
making it into a divisor in $\VZ_{1,(d)}(\PP^m|H,d)$.\sidenote{What about d>m?}

@@

We have produced a factorisation
\begin{equation*}\widetilde\Dcal\to\widetilde \Ecal\to\Ecal,
\end{equation*}
where the first map is flat and the second map is a logarithmic modification. $\widetilde\Dcal$ is a compactification of a torus bundle over $\widetilde\Ecal$ whose strata can be read off from the tropicalization; in particular, we have the information of how (polynomials in) the classes of boundary strata push forward to $\widetilde\Ecal$. In fact, we can say more and describe the generic fibre of $\widetilde\Dcal \to \widetilde\Ecal$:
\begin{lemma}\label{lem:generic_proj_bundle}
Generically over $\Ecal$, the morphism $\widetilde \Dcal\to\Ecal$ is the projectivisation of a tautological vector bundle.
\end{lemma}

@@

Tropically, pushing some of the $\plC_i$ off the circle creates new finite-length edges with coordinates $\ell_i\in\RR_{\geq_0}$ (see the right-hand side of Figure \ref{fig:off_we_go}). Since not all of them can be strictly positive, the resulting fan is that of a projective space. This gives us a fibrewise description of $\widetilde\Dcal\to\Ecal$ over the generic point.

A slight modification of the argument above is required to account for the fact that all vertices that lie on the circle and are mapped to $0\in\mathbb R_{\geq0}$ by the tropical map are generically aligned on $\Ecal$. Indeed, by the combinatorial description of rays in \S \ref{S:combinatorialdescription}, there are no stable vertices inside the circle, so all the edges going from a vertex of $\Lambda_0$ to the core have the same expansion factor $m$ (maximal among the contact orders at the core). Therefore, for all $i$ such that $\plC_i\in\Lambda_0$, the $T_{q_i}C_i^{\otimes m}$ can be identified to the target tangent line on the fibre product \eqref{IIIa fibre product}, and, after saturation, the $T_{q_i}C_i$ themselves are canonically isomorphic on $\Ecal$. Thus, only one summand should be taken to represent all of them in the projective bundle \eqref{eqn:projectivebundle}.
\end{proof}

@@

Consider the locus $\Ecal$ given by $\sigma$ below, and the codimension one locus $\widetilde\Dcal_\rho \subseteq \widetilde\Dcal_\sigma$ given by the two-dimensional (due to the alignment) cone $\rho$:

@@

\begin{figure}[hbt]
\begin{tikzpicture}[ipe import]
  %\useasboundingbox (0, 0) rectangle (595, 842);

@@

\end{tikzpicture}
%\includegraphics[scale=.5]{principalisation}
\caption{On the left, a generic point of $\mathcal P$. On the right, a point of $\Xi$. The dotted circle is the one we draw on $\mathcal P$. The continuous circle represents $\delta$ enlarging the dotted one (a stratum of $\widetilde\Dcal$).}\label{fig:principalisation}
\end{figure}

>>>>

A key step in our recursion is a description of the locus of maps with higher than prescribed tangency, which was identified by Gathmann. We realise the locus as the vanishing locus of a section of a line bundle that comes from tropical geometry — from a piecewise linear function on the fan/tropicalization. The systematic understanding of logarithmic line bundles arising in this fashion is expected to play an important role in logarithmic enumerative geometry.

@@

Our analysis leads naturally to the study of the main component of the double ramification cycle for target manifolds in genus one. The virtual geometry of this was studied in recent work of Janda, Pandharipande, Pixton, and Zvonkine~\cite{DRCBundle}. Even in genus one, an understanding of the main component contributions of the double ramification cycle for target manifolds seems to be lacking. Our main algorithm provides a method for calculating integrals over this class, but it would be more natural to have a formula for it.

@@

The moduli stack $\mathfrak M_{1,n}^{\mathrm{cen}}$  of centrally aligned curves is a logarithmic algebraic stack in the smooth topology, and $\mathfrak M_{1,n}^{\mathrm{cen}}\to\mathfrak M_{1,n}$ is a logarithmic modification. The main construction of~\cite{RSPW} canonically associates to any centrally aligned family of curves $\mathcal C_S$ a partial destabilisation $\widetilde{\mathcal C}_S$, introducing $2$-pointed components, and a contraction
\[
\widetilde{\mathcal C}_S\to \overline{\mathcal C}_S,
\]
where $\overline{\mathcal C}_S$ may contain a Gorenstein elliptic singularity \cite{SMY1}. Intuitively, this is achieved by interpreting the exceptional directions of $\mathfrak M_{1,n}^{\mathrm{cen}}\to\mathfrak M_{1,n}$ as moduli of attaching data.

Formally, given a family of tropical curves, the function $\delta$ provides a non-negative real number for each point on the base. We may therefore consider the ``\textbf{circle}'' of points on each fibre whose distance from the core is equal to $\delta$. A partial destabilization, or ``bubbling'', introduces $2$-pointed rational curves into the curve by adding vertices to the dual graph at the points that lie on this circle (the crosses at height $\delta$ in Figure \ref{fig:lambdaexample}); we denote this new tropical curve by $\widetilde{\plC}$. The open disc of radius $\delta$, i.e. the locus where $\lambda<\delta$, is then contracted to a Gorenstein singularity in $\overline{\mathcal C}_S$. The number of branches is equal to the number of vertices of $\widetilde{\sqC}$ which lie on the circle. %This data uniquely determines the singularity.

@@

The first  factor can be described very concretely as the choice of a point $p$ of $H$, and $d$ lines through it. A very important observation is that, at least generically, the vertices which lie on the circle of radius $\delta$ and map to $0\in\mathbb R_{\geq 0}$ are automatically aligned. Indeed, if $R_i$ denotes the $i$th tail of the curve, glued to $E$ at the node $q_i$, then the stable map provides us with identifications:
\[\operatorname{d}\! f_{i,q_i}\colon T_{R_i,q_i}\cong N_{H/\PP^m,p}\]
which we may compose to obtain attaching data $\theta_{ij}$  (see \S \ref{S:ellsing}). This produces a contraction $C\to\overline C$ to an elliptic $d$-fold point. The dimension of $\widetilde\Dcal$ coincides with the dimension of $\Ecal$, which is $dm + m - 1$.

Assume first that $d\leq m$. For the absolute factorisation to hold, it is necessary (but not sufficient) for the $d$ lines $f_i(R_i)$ to span a subspace of $T_{\PP^m,p}$ of dimension at most $d-1$. The locus of maps satisfying this condition has dimension:
\begin{equation*} (dm+m-1) - (m-d+1) = dm + d -2.\end{equation*}
Without loss of generality, we may assume that $H=\{z_0=0\}$, $p=[0,\ldots,0,1]$, and that the $d$ lines span the subspace $\{z_{d-1}=\ldots=z_{m-1}=0\}$. The sections $z_i$ for $i=d-1,\ldots,m$ vanish identically near $E$, so descend automatically to the singularity. On the other hand the section $z_0$ descends to the singularity by the construction of $\overline C$. Finally, the sections $z_i$ for $i=1,\ldots,d-2$ impose independent conditions to descend to $\overline C$. Imposing these conditions, we find that $\Dcal$ has dimension 
\begin{equation*} (dm+d-2)-(d-2) = dm \end{equation*}
making it into a divisor in $\VZ_{1,(d)}(\PP^m|H,d)$. The case $d>m$ is similar, except that the lines $f_i(R_i)$ are automatically linearly dependent and so this condition does not need to be imposed.

@@

We have produced a factorisation
\begin{equation*}\widetilde\Dcal\to\widetilde \Ecal\to\Ecal,
\end{equation*}
where the first map is flat and the second map is a logarithmic modification. $\widetilde\Dcal$ is a compactification of a torus bundle over $\widetilde\Ecal$ whose strata can be read off from the tropicalization; in particular, we have the information of how (polynomials in) the classes of boundary strata push forward to $\widetilde\Ecal$. In fact, we can say more and describe the generic fibre of $\widetilde\Dcal \to \widetilde\Ecal$:
\begin{lemma}\label{lem:generic_proj_bundle}
Generically over $\widetilde\Ecal$, the morphism $\widetilde \Dcal\to\widetilde\Ecal$ is the projectivisation of a tautological vector bundle.
\end{lemma}

@@

Tropically, pushing some of the $\plC_i$ off the circle creates new finite-length edges with coordinates $\ell_i\in\RR_{\geq_0}$ (see the right-hand side of Figure \ref{fig:off_we_go}). Since not all of them can be strictly positive, the resulting fan is that of a projective space. This gives us a fibrewise description of $\widetilde\Dcal\to\widetilde\Ecal$ over the generic point.

A slight modification of the argument above is required to account for the fact that all vertices that lie on the circle and are mapped to $0\in\mathbb R_{\geq0}$ by the tropical map are already generically aligned on $\widetilde\Ecal$. Indeed, by the combinatorial description of rays in \S \ref{S:combinatorialdescription}, there are no stable vertices inside the circle, so all the edges going from a vertex of $\Lambda_0$ to the core have the same expansion factor $m$ (maximal among the contact orders at the core). Therefore, for all $i$ such that $\plC_i\in\Lambda_0$, the $T_{q_i}C_i^{\otimes m}$ can be identified to the target tangent line on the fibre product \eqref{IIIa fibre product}, and, after saturation, the $T_{q_i}C_i$ themselves are canonically isomorphic on $\widetilde\Ecal$. Thus, only one summand should be taken to represent all of them in the projective bundle \eqref{eqn:projectivebundle}.
\end{proof}

@@

Consider the locus $\Ecal$ given by $\sigma$ below, and the codimension-one locus $\widetilde\Dcal_\rho \subseteq \widetilde\Dcal_\sigma$ given by the two-dimensional (due to the alignment) cone $\rho$:

@@

\begin{figure}[hbt]
\begin{tikzpicture}[ipe import,scale=1.5]
  %\useasboundingbox (0, 0) rectangle (595, 842);

@@

\end{tikzpicture}
%\includegraphics[scale=.5]{principalisation}
\caption{On the left, a generic point of $\mathcal P$. On the right, a point of $\Xi$. The dotted circle is the one we draw on $\mathcal P$. The solid circle represents a radius $\delta$ enlarging the dotted one (a stratum of $\widetilde\Dcal$).}\label{fig:principalisation}
\end{figure}
